
\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{geometry}

\geometry{
 a4paper,
 total={170mm,257mm},
 left=20mm,
 top=20mm,
}

\title{Meeting Minutes: Evaluation Session}
\author{OpenAI's Ada and Mingli}
\date{2023-11-08}

\begin{document}

\maketitle

\section*{Summary}
This meeting primarily focused on setting up the working cycle between OpenAI's Ada and Mingli, assessing capabilities and preferences for the development of Mingli's theory on arithmetic expression geometry.

\section*{Highlights}
\begin{itemize}
    \item Discussed the integration of arithmetic expressions into geometric shapes.
    \item Explored the draft paper on arithmetic expression geometry.
    \item Mingli's requirement for LaTeX document format for math-rich content was noted.
    \item Agreed on the need for both high-level overviews and in-depth technical analysis.
    \item Highlighted the use of rigorous mathematical validation tools like sympy.
    \item Expressed hope for future capability expansions from OpenAI.
\end{itemize}

\section*{Problems}
\begin{itemize}
    \item Current lack of vision capability on Ada's part limits the ability to render LaTeX visuals directly.
    \item No access to external version control systems like Git.
\end{itemize}

\section*{Todos}
\begin{itemize}
    \item Mingli to provide more documents and test cases for further development of the theory.
    \item Ada to continue providing detailed analysis and theory exploration utilizing the existing tools.
    \item Both to remain adaptable to future updates and improvements in Ada's capabilities.
\end{itemize}

\end{document}
