The contact manifold of Section~\ref{sec:ecstructure} supplied the local geometric stage for arithmetic propagation. The present section develops the differential calculus naturally attached to that stage. The guiding idea is simple: the ordinary de Rham differential $d$ differentiates in all directions of $(u,v,a)$, whereas the expression differential $\delta$ keeps only the horizontal part determined by the propagation law. In this sense $\delta$ should be read as a horizontal covariant differential rather than as a second copy of the de Rham operator.

\subsection{Definition and equivalent formulas}

\paragraph{Coordinate definition.}
We define the action of $\delta$ on the coordinate functions by
\begin{equation}\label{eq:9}\tag{9}
\delta a=\omega,
\qquad
\delta u=du,
\qquad
\delta v=dv,
\end{equation}
and extend it to all smooth scalar fields by linearity and the Leibniz rule.

\begin{proposition}\label{prop:delta_equivalent_formula}
For every smooth scalar field $F(u,v,a)$ one has
\begin{equation}\label{eq:10}\tag{10}
\delta F
= dF-(\partial_aF)\,\alpha
= (D_uF)\,du+(D_vF)\,dv.
\end{equation}
Consequently, for every horizontal vector $X\in\ker\alpha$,
\[
\delta F(X)=dF(X).
\]
\end{proposition}

\begin{proof}
Write
\[
dF=(\partial_uF)\,du+(\partial_vF)\,dv+(\partial_aF)\,da.
\]
Since $\alpha=da-(\mu\,du+\lambda a\,dv)$, one gets
\[
dF-(\partial_aF)\alpha
=(\partial_uF+\mu\partial_aF)\,du+(\partial_vF+\lambda a\partial_aF)\,dv,
\]
which is exactly \eqref{eq:10}. If $X\in\ker\alpha$, then $\alpha(X)=0$, so \eqref{eq:10} gives $\delta F(X)=dF(X)$.
\end{proof}

The operators $D_u$ and $D_v$ are therefore the horizontal derivatives singled out by the arithmetic contact structure. In particular, setting $F=a$ in \eqref{eq:10} recovers $\delta a=\omega$, so along any Legendrian curve the equality $\delta a=\omega$ is exactly the propagation law from Section~\ref{sec:flowequation}.

\subsection{Chain rules and elementary examples}

The usual differentiation rules survive, but now with $\delta$ in place of $d$.
For a smooth univariate function $\Phi$ and a smooth bivariate function $H$, one has
\begin{equation}\label{eq:11}\tag{11}
\delta\bigl(\Phi(E)\bigr)=\Phi'(E)\,\delta E,
\qquad
\delta\bigl(H(E_1,E_2)\bigr)=\partial_1H\,\delta E_1+\partial_2H\,\delta E_2.
\end{equation}
The product rule is the special case $H(x,y)=xy$.

Applying \eqref{eq:11} to functions of $a$ alone immediately yields the most useful elementary formulas:
\begin{equation}\label{eq:13}\tag{13}
\delta(a^n)=n a^{n-1}\omega\ \ (n\in\mathbb{Z}),
\qquad
\delta(\ln a)=\frac{\omega}{a}\ \ (a>0),
\qquad
\delta(e^a)=e^a\omega,
\end{equation}
and
\begin{equation}\label{eq:14}\tag{14}
\delta(\sin a)=\cos a\,\omega,
\qquad
\delta(\cos a)=-\sin a\,\omega,
\qquad
\delta(\arcsin a)=\frac{\omega}{\sqrt{1-a^2}}\ \ (|a|<1).
\end{equation}
A particularly useful rectifying example is
\[
y=\arcsinh\!\left(\frac{\lambda a}{\mu}\right)
\qquad\Longrightarrow\qquad
\delta y=\frac{\lambda}{\sqrt{\mu^2+\lambda^2a^2}}\,\omega,
\]
which is the differential form of the rectification already used in Sections~\ref{sec:flowequation} and \ref{sec:01kind}.

For a general mixed field $E=E(u,v,a)$, the horizontal derivatives are
\begin{equation}\label{eq:15}\tag{15}
D_uE=E_u+\mu E_a,
\qquad
D_vE=E_v+\lambda a E_a,
\qquad
\delta E=(D_uE)\,du+(D_vE)\,dv.
\end{equation}
For example, if $E(u,v,a)=ua$, then
\[
D_uE=a+\mu u,
\qquad
D_vE=\lambda ua,
\qquad
\delta E=(a+\mu u)\,du+\lambda ua\,dv.
\]
This is the basic mechanism behind the entire $\delta$-calculus: ordinary derivatives in $(u,v)$ are coupled to the vertical sensitivity $\partial_a$ through the two arithmetic directions.

\subsection{The fundamental curvature package}

For scalar fields we abbreviate the antisymmetrized second horizontal differential by
\[
\delta^2F:=(D_uD_vF-D_vD_uF)\,du\wedge dv.
\]
The main identities can then be packaged as follows.

\begin{theorem}[Fundamental package]\label{thm:delta_fundamental_package}
One has
\begin{equation}\label{eq:12}\tag{12}
[D_u,D_v]=\mu\lambda\,\partial_a,
\qquad
\delta^2F=\mu\lambda(\partial_aF)\,du\wedge dv,
\qquad
d\omega=\mu\lambda\,du\wedge dv+\lambda\,\alpha\wedge dv.
\end{equation}
In particular,
\[
d\omega(D_u,D_v)=\mu\lambda,
\qquad
\delta^2a=\mu\lambda\,du\wedge dv.
\]
\end{theorem}

\begin{proof}
The bracket identity was already computed in Proposition~\ref{prop:local-curvature}:
\[
[D_u,D_v]=\mu\lambda\,\partial_a.
\]
Applying this to a scalar field $F$ gives the middle formula in \eqref{eq:12}. For the last identity, recall that
\[
\omega=\mu\,du+\lambda a\,dv,
\qquad
\alpha=da-\omega.
\]
Hence
\[
d\omega=\lambda\,da\wedge dv
=\lambda(\alpha+\mu\,du+\lambda a\,dv)\wedge dv
=\mu\lambda\,du\wedge dv+\lambda\,\alpha\wedge dv.
\]
Evaluating this on $D_u$ and $D_v$ gives $d\omega(D_u,D_v)=\mu\lambda$, because $\alpha(D_u)=\alpha(D_v)=0$.
\end{proof}

The theorem makes precise the sense in which $\delta$ is not nilpotent. Its square does not vanish; it records the same constant local defect $\mu\lambda$ that already appeared in the contact commutator square of Section~\ref{sec:ecstructure}. Thus $\delta$ is not merely convenient notation for derivatives of expressions. It is a curvature-sensitive differential attached to the arithmetic connection.

\subsection{Connection viewpoint and natural units}

From the bundle viewpoint, the projection
\[
\pi:M=\mathbb{R}^3_{(u,v,a)}\longrightarrow \mathbb{R}^2_{(u,v)},
\qquad
\pi(u,v,a)=(u,v),
\]
together with the horizontal distribution $\mathcal H=\ker\alpha$ defines an Ehresmann connection, and $D_u,D_v$ are the horizontal lifts of $\partial_u,\partial_v$ in that connection language \cite{KobayashiNomizu1963FDG1}. What is special in the present setting is not the abstract notion of a connection, but the arithmetic meaning of its preferred frame: $D_u$ is the infinitesimal add-direction, whereas $D_v$ is the infinitesimal multiply-direction with value-dependent scaling.

It is often convenient to pass to natural units
\[
\tilde u:=\mu u,
\qquad
\tilde v:=\lambda v.
\]
Then
\[
D_u=\mu\bigl(\partial_{\tilde u}+\partial_a\bigr),
\qquad
D_v=\lambda\bigl(\partial_{\tilde v}+a\partial_a\bigr).
\]
This does not change the geometry; it simply removes the dimensional constants from the first-order formulas and makes the computational structure more transparent.

\subsection{A calculable basis and upward sweep}

For explicit calculations, especially in the direction that leads toward function theory, the following scaled monomials are useful. In natural units define
\[
B_n(a,v):=e^{-n\tilde v}a^n
\qquad (n\in\mathbb{N}).
\]
These are adapted to the multiplicative direction because
\[
D_vB_n=0,
\qquad
D_uB_n=\mu n e^{-\tilde v}B_{n-1}.
\]
Thus the family $\{B_n\}$ is invariant along $D_v$ and forms a simple lowering ladder under $D_u$.

To make the closure statement precise, write $q:=e^{\tilde v}$ and consider finite sums
\[
F=\sum_{n=0}^N P_n(u,v,q)\,B_n(a,v),
\]
where each $P_n$ is smooth in $(v,q)$ and polynomial in $u$. Then
\[
D_u\bigl(P_nB_n\bigr)=(\partial_uP_n)B_n+\mu n q^{-1}P_nB_{n-1},
\]
\[
D_v\bigl(P_nB_n\bigr)=\bigl(\partial_vP_n+\lambda q\,\partial_qP_n\bigr)B_n.
\]
So this finite-span class is stable under repeated applications of $D_u$ and $D_v$.

The same basis supports explicit antidifferentiation in the $D_u$-direction. If $P$ is independent of $u$, then
\[
D_u^{-1}\!\bigl(P(v,q)\,B_n\bigr)
=\frac{q\,P(v,q)}{\mu(n+1)}\,B_{n+1}.
\]
More generally, if the coefficients are polynomial in $u$, one removes the remainder by a finite upward sweep, because each correction differentiates the coefficient once in $u$ and therefore lowers its degree.

\paragraph{Example.}
For
\[
F=a^3e^{\lambda v}=e^{4\tilde v}B_3,
\]
one obtains
\[
D_u^{-1}F=\frac{e^{\lambda v}}{4\mu}a^4,
\qquad
D_u\!\left(\frac{e^{\lambda v}}{4\mu}a^4\right)=a^3e^{\lambda v}.
\]
This calculable basis is the main reason the $\delta$-calculus is not merely formal in the present draft: it already supplies explicit families on which the first-order and second-order operators can be computed exactly. The next section uses the same horizontal operators to define the first analytic fields of the theory.
